Group polarization replications may be marred by high false discovery rates

### 2025-06-23 GroupPolarizationStatmod/paper/intro.qmd
worked on Intro

---
title: "If the Null Fits, You Must Omit: Ubiquitous False Detections of Group Polarization"
author: "Matt Turner"
format:
  html:
    toc: true
    include-in-header: header.html
    theme: 
      dark:
        - darkly 
        - theme-dark.scss
      light:
        - theme-light.scss
    fontsize: 0.9em
    linestretch: 1.2
  pdf:
    documentclass: article
    keep-tex: true
bibliography: 
  - this.bib
  - /Users/mt/workspace/Writing/library.bib
output: pdf
---

@@

*Group polarization* is the name for a form of consensus where
members of a like-minded group become, on average, more extreme in their
opinions after discussing a topic. Group polarization is important because
it may increase social and political tensions if moderates become more
extreme. Decades of studies have replicated detections of group
polarization, typically using an ordinal scale to measure opinions.
However, these studies did not account for ordinal measurements, which can
result in spurious detections of opinion change.  Lacking original data we
can still calculate the probability that a published detection of group
polarization was spurious, i.e., to calculate the \emph{false detection
rate}.  Across 54 group polarization experimental conditions from ten
representative journal articles, we calculated false detection rates to be
between 0.52 and 0.88, with a median of 0.75, using a generative model of
group polarization experiments seeded with empirical data.  
We also use our model to develop experimental designs that
achieve an acceptable false discovery rate.  Much of group polarization
research may be unreliable. This analysis can help change that by enabling others to avoid this
hidden pitfall. More
broadly this work demonstrates one important way that replication success alone
does not imply epistemic reliability.

@@

> A medium effect size is conceived as one large enough to be visible to the naked eye. [@CohenBook1988]

@@

@@

**🧭 Next session, 6/20: Map paragraph set bridging theory to methods — experiments → ordinal pitfalls → null-consistency model → power analysis → transition to results**

## Bridge: From Philosophy to Mechanics {.unnumbered}

<!-- 🧭 Writing plan for next session -->

@@

- Brief overview of how group polarization experiments are typically designed  
- Emphasize reliance on **pre/post ordinal opinion measures**  
- Highlight the assumption that ordinal differences reflect latent opinion shifts

@@

- Explain why ordinal measurements do **not** uniquely reflect latent shifts  
- Use an example or intuition (e.g., compressing/shrinking variance without changing means)  
- Show how this opens the door to *null-consistent but polarization-looking* results

@@

- Introduce the generative model with fixed latent mean and changing variance  
- Explain what counts as a plausible fit and how the model reverse-engineers the observation  
- Clarify that this is a *constructive falsification tool*, not a probabilistic inference

@@

- Discuss how, even if the detection is real, underpowered studies can't distinguish it from null  
- Mention use of Bayesian ordered probit models to test for minimum detectable effect sizes  
- Preview that even with ideal modeling, many studies likely lack sufficient power

@@

- Signal that the next section tests this logic across all experimental conditions  
- Frame the upcoming analysis as assessing the **robustness** of published detections under this new lens

>>>>

---
title: "If the Null Fits, You Must Omit: Ubiquitous False Detections of Group Polarization"
author: "Matt Turner"
format:
  html:
    toc: true
    include-in-header: header.html
    theme: 
      dark:
        - darkly 
        - theme-dark.scss
      light:
        - theme-light.scss
    fontsize: 0.8em
    linestretch: 1.1
  pdf:
    documentclass: article
    keep-tex: true
bibliography: 
  - this.bib
  - /Users/mt/workspace/Writing/library.bib
output: pdf
---

@@

*Group polarization* is the name for a form of consensus where
members of a like-minded group become, on average, more extreme in their
opinions after discussing a topic. Group polarization is important because
it may increase social and political tensions if moderates become more
extreme. Decades of studies have replicated detections of group
polarization, typically using an ordinal scale to measure opinions.
However, these studies did not account for ordinal measurements, which can
result in spurious detections of opinion change.  Lacking original data we
can still calculate the probability that a published detection of group
polarization was spurious, i.e., to calculate the *false detection
rate*.  Across 54 group polarization experimental conditions from ten
representative journal articles, we calculated false detection rates to be
between 0.52 and 0.88, with a median of 0.75, using a generative model of
group polarization experiments seeded with empirical data. We also use our model to develop experimental designs that achieve an acceptable false discovery rate.  Much of group polarization research may be unreliable. This analysis can help change that by enabling others to avoid this hidden pitfall. More broadly this work demonstrates one important way that replication success alone
does not imply epistemic reliability.

@@

<!-- > A medium effect size is conceived as one large enough to be visible to the naked eye. [@CohenBook1988] -->

@@

- We demonstrate here that there is a pervasive problem observed over decades of group polarization research where faulty statistical methods and incomplete reporting that should force the retraction of these studies from the canon. Effectively, then, empirical support for the theoretical concept *group polarization* stops being so: the epistemic integrity of *group polarization* is undermined with its empirical foundation losing mass.

We complement this critical analysis with a prescriptive analysis to demonstrate that more appropriate statistical models do indeed lower the false discovery rate by committing fewer false positive detections of group polarization under a null model with shrinking opinion variance. Recently, statistical best-practice is to use Bayesian models of statistics, where social dynamics and opinion-reporting behaviors are represented as causally-dependent probability distributions, where a sample taken from one distribution (e.g., the opinion distribution of a group in a group polarization experiment) acts as an input parameter to another distribution (e.g., that represents the individually-heterogeneous distortion of opinions when reported on an ordinal scale). Our models follow current best practices in psychometrics, social science, and Bayesian inference [@Liddell2018; @Kruschke2015; @Rethinking; @Deffner2024], but consolidate them in a new, useful way that enables the
retroactive discovery of null model parameters that generate spurious group polarization, forcing published evidence for group polarization to be classified as non-identifiable, and removed from evidence that can be said to support group polarization.

Our methods can be turned on any data-driven social science analysis of data where participants report a latent, subjective psychological variable (mood, physiological state, etc.), and where various theories supposedly battle it out to best represent the phenomenon of interest via NHST performed on ordinal observations with weak, informal theoretical models to motivate the studies. This may just be the first of many bricks to fall since *all 68* of the articles that mentioned the word "Likert" and analyzed ordinal data used metric models to detect latent psychological variable change across three influential psychology journals, *Psychological Science*, the *Journal of Experimental Psychology: General*, and the *Journal of Personality and Social Psychology* [@Liddell2018].

@@

## Identifiability and the Illusion of Change

<!-- XXX BULLET POINTS ALL GENERATED BY CHATGPT BASED ON CONVO TO HELP 
     MAKE A SNAPSHOT OF ALL THE KEY POINTS THAT I HAVE DEVELOPED INDEPENDENTLY
     BUT AM STRUGGLING TO HOLD TOGETHER IN MY HEAD AT ONCE. I WILL BE EDITING, 
     CUSTOMIZING, ETC., AFTER WHICH THIS MESSAGE WILL BE REMOVED. 6/23/25 -->

- Many group polarization studies report opinion change based on pre/post differences, but assume a transparent mapping between observed scores and latent beliefs.
- Yet ordinal opinion data can mask deeper structural ambiguities that invalidate these assumptions.
- In what follows, we first describe the basic structure of group polarization experiments, then explain how ordinal measurement can lead to illusory shifts.
- Finally, we introduce a null-consistency modeling procedure that operationalizes this insight, and show how it informs the credibility of reported effects.

@@

- Most group polarization experiments follow a familiar structure: measure individual opinions, allow group discussion, then re-measure.
- The central claim is that like-minded individuals become more extreme after deliberation.
- Crucially, opinions are typically measured with Likert-style ordinal scales (e.g., from “Strongly disagree” to “Strongly agree”).
- Analyses often treat these as interval-scaled, calculating mean differences and testing for significance.

@@

- Ordinal scores do not have a fixed interval structure — they represent ranked categories, not true distances.
- A given change in score may reflect a shift in opinion, increased certainty, or both.
- For example, if opinions become more tightly clustered (lower variance) post-deliberation, the group mean can appear to shift even when the latent mean remains unchanged.
- This opens the door to *null-consistent detections* — apparent effects that emerge from measurement artifacts rather than real opinion movement.

@@

- We formalize this concern using a generative model that holds the latent opinion mean constant while allowing variance to shrink.
- The goal is to test whether an observed group polarization result could arise from such a stable-mean, shifting-variance process.
- If this model can reproduce the pre/post measurement pattern, the observed result is non-diagnostic — it is consistent with a null hypothesis of no true change.
- This is a form of constructive falsification: it doesn’t prove the effect is false, but shows it is not uniquely explained by change.

@@

- Even if an effect is real, detecting it reliably requires statistical power — the ability to distinguish true shifts from noise or artifacts.
- Studies using ordinal measures and small samples may be especially vulnerable to false positives under the null.
- To assess this, we fit Bayesian ordered probit models to simulated data and estimate the power to detect effects of various sizes.

@@

- Having established a model-based definition of null-consistent detection, we now apply it to published group polarization experiments.
- By combining this logic with simulations seeded from published data, we estimate false discovery rates across conditions.
- This isn’t a judgment about whether past detections were true. It’s a demonstration of how to estimate false detection rates properly—using models that actually fit the structure of the data. We calculate FWER and FDR for the original study designs at three effect size levels—small, medium, and large—based on Cohen’s guidelines [@CohenBook1988, Ch. 2.2.3] and Bayesian methods from Liddell and Kruschke [@Liddell2018]. The verdict is clear: even if these studies had used better models, most still wouldn’t have had the power to rule out false positives. That makes replication a weak signal of reliability. If future group polarization studies want to produce meaningful results, they need stronger designs—and that starts with a clean break from outdated statistical habits.

### 2025-06-23 GroupPolarizationStatmod/paper/ms.tex
worked on Intro

This work is motivated by a core conviction that better thinking leads to better
science. Theoretical clarity, careful measurement, and principled statistical modeling are not academic luxuries, but essential tools for empirical science. The group polarization literature has long suffered from ambiguities in theory, ordinal-to-continuous mismatches in measurement, and default inferential techniques that mask model non-identifiability. By reconstructing these issues from the ground up, this paper aims not only to assess the credibility of existing evidence, but to show that rigorous thinking — about what we measure, how we model, and what we can legitimately conclude — has direct, practical consequences for the claims we make about the social world.

>>>>

This work is motivated by a core conviction that better thinking leads to better
science. Theoretical clarity, careful measurement, and principled statistical
modeling are not academic luxuries, but essential tools for empirical science.
Theoretical and statistical confusions in the group polarization literature have
lain dormant, undermining measurements of opinion change in group polarization
unbeknownst to generations of researchers~\autocite{Brown1986}.

literature has long suffered from ambiguities in theory, ordinal-to-continuous mismatches in measurement, and default inferential techniques that mask model non-identifiability. By reconstructing these issues from the ground up, this paper aims not only to assess the credibility of existing evidence, but to show that rigorous thinking — about what we measure, how we model, and what we can legitimately conclude — has direct, practical consequences for the claims we make about the social world.
